We have presented two classes of pattern detection for the formulas with
multipliers implemented using long school multiplier and Wallace tree.
%
We have implemented the algorithms and applied to several handcrafted and 
industrial benchmarks.
%
The experiments suggest a significant improvement in performance.

We are currently extending our procedure to support Booth multiplier
and other difficult arithmetic patterns.
%
We are also working to add proof generation support
for the added tautologies.
%
We could not include proof generation in this work, since basic infrastructure
of proof generation is missing in $\zthree$~bit-vector rewriter module.
%
We are also planning to request \zthree~team to adopt our modification in their
standard distribution.


We plan to extend our work to problems involving different combinations of multiplications. For example, $(X*(Y*Z))*W$ involves three multiplications. We could chose to specify the implementation of the three multiplications to be either of Long, Wallace, Booth multiplication or an unspecified multiplication. To enable the pattern detection in arbitrary arithmetic formulas, we are working to modify the word level reasoning in the solvers.  
%we could further extend our work to complicated word level formulas involving bit vector multiplication like $X*X + Y*Y$, $X*(Y+Z)$, $(X*(Y*Z))*W$.

%None of the solvers were able to make use of the added tautologies(one for each long multiplication). 
%The problem seems to be on two fronts: one is that to check equality of two terms, the solvers want exact structural similarity and secondly the infrastructure of using one assertion to infer the satisfiability of another does not seem to be well developed. 
%We aim to bridge this gap as part of our future work.



%--------------------- DO NOT ERASE BELOW THIS LINE --------------------------

%%% Local Variables: 
%%% mode: latex
%%% TeX-master: "main"
%%% End: 
