%
We have implemented\footnote{\url{https://github.com/rahuljain1989/Bit-vector-multiplication-pattern}} our algorithms as a part of $\zthree$ SMT solver.
%
We evaluate the performance of our algorithms using benchmarks that are industrial and handcrafted hardware verification problems.
%
We compare our tool with $\zthree$, $\boolector$ and $\cvcfour$.
%
Our experiments show that the solvers time out on most of the benchmarks and our tool produces results within the set time limit.

\paragraph{\bf Implementation}
We have added about 1500 lines of code in the bit vector rewrite module of $\zthree$ because it allows an easy access to the abstract syntax tree of the input formula. We call this version of $\zthree$ as instrumented-$\zthree$.
%
An important aspect of the implementation is the ability to exit as early as possible if the match is going to fail.
%
We implemented various preliminary checks including the ones mentioned in Algorithm 1. For example, we ensure that the size of $\Lambda_i$ is upper bounded appropriately as per the scheme of long multiplication. We exit as soon as the upper bound is violated. 
%
We have implemented three versions of $\ourtool$ by varying the choice of $\textsc{SMTSolver}$.
%
We used  $\zthree$, $\boolector$, and $\cvcfour$ for the variations. 
%

In each case we stop the instrumented-$\zthree$ solver after running our matching algorithms,
print the learned tautologies in a file along with the input formula, and
run the solvers in a separate process on the pre-processed formula. 
The time taken to run our matching algorithms and generate the pre-processed formula is less than one second across all our benchmarks, and hence is not reported. 
We also experimented by running instrumented-$\zthree$ standalone and found the run times to be similar to the above $\zthree$ case; hence the run times for instrumented-$\zthree$ are not reported. We use the following versions of the solvers: $\zthree$(4.4.2), $\boolector$(2.2.0), $\cvcfour$(1.4).

\paragraph{\bf Benchmarks}
%
Our experiments include 20 benchmarks. 
%
Initially, we received the motivating example described in Section~\ref{sec:intro} involving long multiplication that was not solved by any
of the solvers in 24 hours.
%
This example inspired our current work and to evaluate it we generated several similar benchmarks.
%
For long multiplication, we generated benchmarks by varying three characteristics, firstly the total bit length of the input bit-vectors, secondly the width of each block, and thirdly assigning specific blocks as equal or setting them to zero.
%
Our benchmarks were written in $\sysver$ and fed to STEWord~\cite{wste}, a hardware verification tool.
%
STEWord takes $\sysver$ design as input and generates the corresponding SMT1 formula.
%
We convert the SMT1 formula to SMT2 format using $\boolector$.
%
In the process, $\boolector$ extensively simplifies the input formula but retains the overall structure.
%
We have generated benchmarks also for Wallace tree multiplier similar to the long multiplication.
%
For $n$-bit Wallace tree multiplier, we have written a script that takes $n$ as input and generates all the files needed as input by STEWord. All our benchmarks correspond to the system implementation satisfying the specified property: in other words, the generated SMT formulas were unsatisfiable. For satisfiable formulas the solver was able to find satisfying assignments relatively quickly, both with and without our heuristic. Hence, we do not report results on satisfiable formulas.
%

\paragraph{\bf Results}
%
We compare our tool with $\zthree$, $\boolector$ and $\cvcfour$.
%
In the Tables~\ref{tbl:time}-\ref{tbl:cd}, we present the results of the experiments.
%
We chose timeout to be 3600 seconds.
%
In Table~\ref{tbl:time}, we present the timings of the long multiplication and
Wallace tree multiplier experiments.
%
The first 13 rows correspond to
the long multiplication experiments.
%
The columns under $\textsc{SMTSolver}$ are the run times of the
solvers to prove the satisfiability of the input benchmark.
%
The solvers timeout on most of the benchmarks.
%

\input{tbl-long-time}

The next three columns present the run times of the three versions of
$\ourtool$ to prove the satisfiability of the benchmarks.
%
$\ourtool$ with $\cvcfour$ makes best use of the added tautologies.
%
$\cvcfour$ is quickly able to infer that the input formula and the
added tautologies are negations of each other justifying the timings.
%
$\ourtool$ with $\boolector$ and $\zthree$ does not make the above
inference, leading to more running times.
%
$\boolector$ and $\zthree$ bit blast the benchmarks having not been
able to detect the structural similarity.
%
However, the added tautologies help $\boolector$ and $\zthree$ to
reduce the search space, after the SAT solver is invoked on the bit
blasted formula.


The last 7 rows correspond to the Wallace tree multiplier experiments.
%
Since the multiplier involves a series of half and full adders, the
size of the input formula increases rapidly as the bit vector size
increases.
%
Despite the blowup in the formula size, $\ourtool$ with $\zthree$ is
quickly able to infer that the input formula and the added tautology
are negations of each other.
%
However, $\ourtool$ with $\boolector$ and $\cvcfour$ do not make the
inference, leading to larger run times.
%
This is because of the syntactic structure of the learned tautology
from our implementation inside $\zthree$.
%
The input formula has `and' and `not' gates as its building blocks,
whereas $\zthree$ transforms all `ands' to `ors'.
%
Therefore, the added tautology has no `ands'.
%
The difference in the syntactic structure between the input formula
and the added tautology makes it difficult for $\boolector$ and
$\cvcfour$ to make the above inference.
%

We have seen that the solvers fail to apply the word level reasoning
after adding the tautologies.
%
In the case, the solvers bit blast the formula and run a SAT solver.
%
In Table~\ref{tbl:cd}, we present the number of conflicts and decisions within
the SAT solvers.
%
The number of conflicts and decisions on running $\ourtool$ with the
three solvers, are considerably less than their $\textsc{SMTSolver}$
counterparts in the most of the cases.
%
This demonstrates that the tautologies also help in reducing the
search inside the SAT solvers. 
%
$\ourtool$ with $\cvcfour$ has zero conflicts and
decisions for all the long multiplication experiments, because the 
word level reasoning solved the benchmarks.
%
Similarly, $\ourtool$ with $\zthree$ has zero conflicts and decisions
for all the Wallace tree multiplier experiments.

% \ashu{
% %
% We also applied $\ourtool$ with $\zthree$ on SMT-LIB benchmarks and compared 
% with the original $\zthree$.
% %
% $\ourtool$ was able to handle $..\%$ of the benchmarks.
% %
% We observed timings for $..\%$ of the benchmarks did not change more than $5\%$,
% $..\%$ of the benchmarks increased more than $5\%$,
% and 
% $..\%$ of the benchmarks decreased more than $5\%$.
% }

\input{tbl-long-cd}
%

\paragraph{\bf Limitations}
%
Our implementation and method have several limitations
in solving the target bit-vector problems.
%
We have only considered a limited space of implementations of the
multipliers and the implementations may include several other
optimizations, e.g., multiplying with constants using bit-shifting.
%
The multipliers may be applied recursively, e.g.,
the partial products of a long multiplication may be
obtained using Wallace tree multiplier.
%
Adding assertions in the formula may hurt solving time, because
it may also be bit-blasted by the solver and overwhelms
the underlying SAT solver. 
%
On top of that, the added assertions themselves are rewritten and
unrecognizable by the equality checker.
%
Since the nature of our method is to exploit the potential
structure in the input, we must also adopt all parts of the solver
to be aware of the sought structure, which is our future goal.
%
We are currently working to tag the added assertions such that they
are neither simplified in pre-processing nor bit-blasted in the final
solving.
%
They should only contribute to the word-level reasoning.
%
Our experiments do not include examples where multiplication
is embedded deep in a large formula.
%
We are also working to make our implementation robust such that
it can reliably work on larger examples, in particular on all the 
SMT-LIB benchmarks.
%
More results may be found in the direction at~\cite{arxiv-version}.


%--------------------- DO NOT ERASE BELOW THIS LINE --------------------------

%%% Local Variables: 
%%% mode: latex
%%% TeX-master: "main"
%%% End:




