% Bit-vector is important
%
In recent years, SMT solving has emerged as a powerful technique for
testing, analysis and verification of hardware and software systems.
A wide variety of tools today use SMT solvers as part of their core
reasoning
engines~\cite{hwcbmc,boolector,ebmc,cbmc,corral,boogie,crv1,crv2,dart,concolic}.
%; examples include bounded model
%checkers~\cite{hwcbmc,boolector,ebmc,cbmc}, static assertion
%checkers~\cite{corral,boogie}, word-level symbolic trajectory
%evaluators~\cite{wste}, constrained test
%generators~\cite{crv1,crv2,dart}, concolic simulators~\cite{concolic},
%among others.
A common approach used in several of these tools is to model the
behaviour of a system using formulas in a combination of first-order
theories, and reduce the given problem to checking the
(un)satisfiability of a formula in the combined theory.  SMT solvers
play a central role in this approach, since they combine decision
procedures of individual first-order theories to check the
satisfiability of a formula in the combined theory.  Not surprisingly,
heuristic techniques to improve the performance of SMT solvers have
attracted significant attention over the years %% .  The literature
%contains a
%rich
%% body of heuristic strategies for improving the performance of
%% theory-specific solvers
(see~\cite{barrett,deMoura2013} for excellent expositions).  In this
paper, we add to the repertoire of such heuristics by proposing a
pre-processing step that analyzes an input formula, and adds specially
constructed assertions to it, without changing the semantics. We focus
on formulas in the quantifier-free theory of fixed-width bit-vectors
with multiplication, and show by means of experiments that three
state-of-the-art SMT solvers benefit significantly from our heuristic
when solving many benchmarks with multiplication operators.%% , our
%heuristic can significantly reduce the
%% solving time yields significant performance benefits in many cases for
%% namely {\zthree}~\cite{zthree}, {\cvcfour}~\cite{cvcfour} and
%% {\boolector}~\cite{boolector}.
%% %Significantly, our heuristic helps reduce the solving time for
%multiple examples by upto several orders of magnitude when the input
%formula is unsatisfiable.

%% Theories commonly
%% supported in modern SMT solvers include the quantifier-free theories
%% of fixed-width bit-vectors, arrays, lists, strings, among
%% others~\cite{kroening-book,smtlibv2}.
%% Since reasoning in these individual theories is often much more
%% efficient than reasoning about the bit-level representation of the
%% corresponding data types, techniques based on SMT solving hold a lot
%% of promise as far as scaling to large applications is concerned.  The
%% impressive progress made in SMT solving over the last two
%% decades~\cite{smtprogress} has also substantially lived up to this
%% hope. 

The primary motivation for our work comes from word-level bounded
model checking (WBMC)~\cite{cbmc,hwcbmc} and word-level symbolic
trajectory evaluation (WSTE)~\cite{wste} of embedded hardware systems.
Specifically, we focus on systems that process data, represented as
fixed-width bit-vectors, using arithmetic operators.
%% Examples of such
%% systems include digital signal processing filters, graphics
%% accelerators, encryption and decryption modules, custom datapath
%% implementations etc.  
When reasoning about such systems, it is often necessary to check
whether a high-level property, specified using bit-vector arithmetic
operators (viz. addition, multiplication, division), is satisfied by a
model of the system implementing a data-processing algorithm.  For
reasons related to performance, power, area, ease of design etc.,
complex arithmetic operators with large bit-widths are often
implemented by composing several smaller, simpler and
well-characterized blocks.  For example, a $128$-bit multiplier may be
implemented using one of several multiplication
algorithms~\cite{long,booth,wallace}
%viz. long multiplication~\cite{long}, Booth-encoded
%multiplication~\cite{booth} or Wallace-tree
%multiplication~\cite{wallace},
after partitioning its $128$-bit operands into narrower blocks.  SMT
formulas resulting from WBMC/WSTE of such systems are therefore likely
to contain terms with higher-level arithmetic operators
(viz. $128$-bit multiplication) encoding the specification, and terms
that encode a lower-level implementation of these operators in the
system (viz. a Wallace-tree multiplier).  Efficiently reasoning about
such formulas requires exploiting the semantic equivalence of these
alternative representations of arithmetic operators.  Unfortunately,
our study, which focuses on systems using the multiplication operator,
reveals that three state-of-the-art SMT solvers
({\zthree}~\cite{zthree}, {\cvcfour}~\cite{cvcfour} and
{\boolector}~\cite{boolector}) encounter serious performance
bottlenecks in identifying these equivalences.  Our limited
experiments show that these bottlenecks manifest most conspicuously
when reasoning about the unsatisfiability of formulas.

\noindent {\bfseries \emph{A motivating example:}} To illustrate
the severity of the problem, we consider the SMT formula arising out
of WSTE applied to a pipelined serial multiplier circuit, originally
used as a benchmark in~\cite{wste}.  The circuit reads in two $32$-bit
operands sequentially from a single $32$-bit input port, multiplies
them and makes the $64$-bit result available in an output register.
%% The circuit also has
%% several control signals that can be used to change the flow of
%% control, effectively delaying the computation of the result.

The property to be checked asserts that if $a$ and $b$ denote the
word-level operands that are read in, then after the computation is
over, the output register indeed has the product $a *_{[32]} b$, where
$*_{[32]}$ denotes $32$-bit multiplication.  The system implementation,
as used in~\cite{wste}, is described in $\sysver$ (a hardware description
language) and makes use of the multiplication operator (i.e., {\tt *}) in
$\sysver$ with $32$-bit operands.  The Language Reference Manual of
$\sysver$ specifies that this amounts to using a $32$-bit
multiplication operation directly.  The SMT formula resulting from a
WSTE run on this example therefore contains terms with only $32$-bit
multiplication operators, and no terms encoding a lower-level
multiplier implementation.  This formula is proved unsatisfiable
in a fraction of a second by {\boolector}, {\cvcfour}
and {\zthree}.%% Note that in WSTE (as also in WBMC), the SMT formula
%% encodes violation of a property by a bounded run of the system. Hence,
%% unsatisfiability of the formula implies the absence of any bounded
%% violating runs.

We now change the design above to reflect the implementation of
$32$-bit multiplication by the long-multiplication
algorithm~\cite{long}, where each $32$-bit operand is partitioned into
$8$-bit blocks.  The corresponding WSTE run yields an SMT formula that
contains terms with $32$-bit multiplication operator (derived from the
property being checked), and also terms that encode the implementation
of a $32$-bit multiplier using long-multiplication.  Surprisingly,
none of {\boolector}, {\cvcfour} and {\zthree} succeeded in deciding
the satisfiability of the resulting formula even after $24$ hours on
the same computing platform as in the original experiment.  The
heuristic strategies in these solvers failed to identify the
equivalence of terms encoding alternative representations of $32$-bit
multiplication, and proceeded to bit-blast the formulas, resulting in this
dramatic blowup in run-time.

\noindent {\bfseries \emph{Problem formulation:}} 
The above example demonstrates that the inability to identify semantic
equivalence of alternative representations of arithmetic operators
plagues multiple state-of-the-art SMT solvers.  %% Therefore, a heuristic
%% that helps in this respect and is generic (not solver-specific) would
%% be highly desirable.
This motivates us to ask: \emph{Can we heuristically pre-process an
SMT formula containing terms encoding alternative representations of
bit-vector arithmetic operators, in a solver-independent manner, so
that multiple solvers benefit from it?}  We answer this question
positively in this paper, for the multiplication operator.  The
motivating example, that originally timed out after $24$ hours on
three solvers, is shown to be unsatisfiable by {\zthree} in 0.073s and
by {\cvcfour} in 0.017s, after applying our heuristic. Although
{\boolector} does not benefit from our heuristic on this example,
it benefits in several other examples, as discussed in
Section~\ref{sec:experiments}.

\noindent {\bfseries \emph{Term re-writing vs adding tautological assertions:}} 
Prima facie, the above problem can be solved by reverse-engineering a
lower-level representation of a bit-vector arithmetic operator, and by
re-writing terms encoding this representation with terms using the
higher-level bit-vector operator.  Indeed, variants of this approach
have been used earlier in different
contexts~\cite{kunz,ciesielski,kolbl,reveng,earlier-pat-match-synopsys}.
In the context of SMT solving, however, more caution is needed.
%we need to be careful%% complications can potentially
%% arise if we simply
%before re-writing a term encoding one representation of an arithmetic
%operator by another term encoding a different representation.
As shown in Example $2$ of Section~\ref{sec:long-mult}, the same
collection of terms (in this case, sums-of-partial-products) can arise
from two different word-level multiplication operations.  This makes it
difficult to decide which of several term re-writes should be
used when there are alternatives. %% to help the SMT solver
%% decide the satisfiability of the input formula.
%Furthermore,
Even if the above dilemma doesn't arise, re-writing one term with
another is a ``peep-hole'' transformation that may not always
correlate with improved solver performance. % for various
%reasons.  %% Indeed,
%% term re-writing is a ``peep-hole'' transformation that is oblivious of
%% the overall context in which the terms appear in the SMT formula.
%% What appears beneficial locally may not be beneficial in the overall
%% (un)satisfiability check.  In addition,
%% For example, syntactially distinct terms that are semantically
%% equivalent may play different roles when reasoning about different
%% sub-formulas of an SMT formula. 
For example, one term may enable a re-write rule that helps simplify
one sub-formula, while a syntactically distinct but semantically
equivalent term may enable another re-write rule that helps simplify
another sub-formula. Re-writing one term by another prevents
both terms from jointly contributing to simplifications and improving
the solver's performance.%satisfiability check.

%% Re-writing a term encoding a low-level representation of an arithmetic
%% operator with another term representing the same operator at a higher
%% level may not always benefit an SMT solver.  
In this paper, we propose a heuristic alternative to term re-writing
when solving bit-vector formulas with multiplication.  Given a
bit-vector formula $\varphi$ containing terms with different
representations of multiplication, our heuristic searches for patterns
in the terms corresponding to two multiplication algorithms, i.e.,
long multiplication and Wallace-tree multiplication. Instead of
re-writing the matched terms directly with bit-vector multiplication
terms, we conjoin $\varphi$ with assertions that semantically equate a
matched term with the corresponding bit-vector multiplication term.
Note that each added assertion is a tautology, and hence does not
change the semantics of the formula.  Since no re-writes are done, we
can express multiple semantic equivalences without removing any
syntactic term from the formula.  This is an important departure from
earlier techniques, such as~\cite{kolbl}, that rely on sophisticated
re-writes of the formula. Our experiments show that the added
tautological assertions succeed in preventing bit-blasting while
solving in several cases, while in other cases, they help in pruning
the search space even after bit-blasting.  Both effects eventually
translate to improved performance of the SMT solver.  Furthermore,
since our heuristic simply adds assertions to the input formula, it is
relatively independent of the internals of any specific solver, and
can be used with multiple solvers. %% Our
%experiments show that the
%performance
%% of different SMT solvers on the pre-processed formulas can vary.
%% Hence, we propose a portfolio approach to solving the pre-processed
%% formulas.  We show experimentally that a portfolio solver using
%% pre-processed formulas significantly outperforms a portfolio solver
%% using the original formulas.



%% Many hardware and software verification problems are translated to the
%% satisfiability of quantifier free bit-vector(QF\_BV)
%% formulas~\cite{hardware,cbmc,more}.
%
