%
In this section, we will present our method for solving
formulas that contain implementations of multiplications.
%
Our method first attempts to match multiplications that
are decomposed using long or Wallace tree multiplication.
%
If we match some subterms of the input formula
are indeed instances of the
multiplications, we add tautologies stating that the terms are 
equal to the multiplications of the matched bit-vectors.
%
Our matching method may find multiple matches for a subterm.
%
We add a tautology for each match.
%
Let us first present our method of matching long multiplication. 

\subsection{Matching long multiplication}
\input{alg-match-long}

In Algorithm~\ref{alg:long}, we present a function $\textsc{MatchLong}$
that takes a $\qfbv$ term $t$ and returns a set of matched multiplications.
%
The algorithm and the subsequent algorithms are written such that as soon
as it becomes clear that no multiplication can be matched then
they return empty set. 
%
At line 1, we match $t$ with a sum of concatenations and if the match fails
then clearly $t$ is not a long multiplication.
%
At line 2, we find a partial product among $s_{ij}$ and extract
the block size $w$ used by the long multiplication.
%
The loop at line $4$ populates the vector of the set of partial products $\Lambda$.
%
$\Lambda_i$ denotes the partial products that are aligned at the $i$th block.
%
Each $s_{ij}$ must either be $0$ or a partial product of width $w$.
%
Otherwise, $t$ is declared unmatched at line 8. 
%
At line 5, we compute the alignment $o$ for $s_{ij}$.
%
If $s_{ij}$ happens to be a partial product then it is inserted in
$\Lambda_o$ at line 7.
%
At line 9, we call $\textsc{getMultOperands}$ to identify the operands
of the multiplication from $\Lambda$ if $t$ is indeed a long
multiplication.

\subsection{Partial products to operands}
\input{alg-get-operand}

In Algorithm~\ref{alg:operand}, we present a function
$\textsc{getMultOperands}$ that takes a vector of multiset of partial
products $\Lambda$ and block width $w$, and returns a set of matched
multiplications.
%
The algorithm proceeds by choosing a pair of operands with
insufficient information and backtracks if the guess is found to be
wrong.


At line 1, we compute $h$ and $l$ that establishes the range of search for
the operands.
%
We maintain two candidate operands $x$ and $y$ of size $hw$.
%
We also maintain a vector of bits $backtrack$ that encodes 
the possibility of flipping the uncertain decisions.
%
Due to the scheme of the long multiplication, the highest
non-empty entry in $\Lambda$ must be a singleton set.
%
If $\Lambda_h$ contains a single partial product $a*b$,
we assign $x_h$ and $y_h$ the operands of $a*b$ arbitrarily.
%
Otherwise, we declare the match has failed by returning $\emptyset$.
%
We assign $\lfalse$ to $backtrack_h$, which states that
no need of backtracking at index $h$.
%
The loop at line 8 iterates over index $i$ starting from $h$ and decreases
$i$ in each iteration.
%
In each iteration, it assigns values to $x_i$, $y_i$, and $backtrack_i$. 
%

The algorithm may not have enough information at the $i$th iteration
and the chosen value for $x_i$ and $y_i$ is be wrong.
%
Whenever, the algorithm realizes that such a mistake has happened
it jumps to line 31.
%
It increases back the value of $i$ to the latest $i'$ that allows
backtracking.
%
It swaps the assigned values of $x_i$ and $y_i$, and disables future
backtracking to $i$ by setting $backtrack_i$ to
$\lfalse$.
%

Let us look at the loop at line 8 again.
%
We also have variables $l_x$ and $l_y$ that contain the index of
the least non-zero entries in $x$ and $y$, respectively.
%
At line 9, we decrement $i$ and $\Lambda_i$ is copied to $C$.
%
At index $i$, the sum of the aligned partial products is the following.
$$
x_{h}*y_{i} + \underbrace{x_{h-1}*y_{i+1} + \dots + x_{i+1}*y_{h-1}}_{\text{operands seen at the earlier iterations}} + x_{i}*y_{h}
$$
We have already chosen the operands of the middle multiplications in
the previous iterations.
%
Only the multiplications at the extreme ends have $y_i$ and $x_i$ that are
not assigned yet.
%
In the loop at line 10, we remove the middle multiplications.
%
If any of the needed multiplication is missing then we may have made a mistake
earlier and we jump for backtracking.
%
After the loop, we should be left with at most two partial products in $C$.
%
We match $C$ with five patterns at lines 14-19 and
update $x_i$, $y_i$, and $backtrack_i$ accordingly.
%
If none of the pattern match, we jump for backtracking at line 20.
%
In some cases we clearly determine the value of $x_i$ and $y_i$, and
we are not certain in the other cases.
%
We set $backtrack_i$ to $\ltrue$ in the uncertain cases to indicate
that we may return back to index $i$ and swap $x_i$ and $y_i$.
%
In the following list, we discuss the uncertain cases.
%
\begin{itemize}
\item[line 15:] If $C$ has two elements $x_h*b$ and $y_h*d$,
there is an ambiguity in choosing $x_i$ and $y_i$
if $x_h == y_h$.
%
% Thus, if  $x_h == y_h$, we set $backtrack_i$ to $\ltrue$.
\item[line 16:] If $C$ has a single element $x_h*y_h$, there  
are two possibilities.
\item[line 17:] If $C = \{x_h*b\}$ and $b$ is not $y_h$ then 
  similar to the first case there is an ambiguity in
  choosing $x_i$ and $y_i$ if $x_h == y_h$. Line 18 is similar.
\item[line 19:] If $C$ is empty then there is no uncertainly. % $x_i = y_i = 0$.
\end{itemize}
%
At line 21-22, we update $l_x$ and $l_y$ appropriately.
%
The condition at line 23 ensures that the expected
least index $i$ such that $\Lambda_i \neq \emptyset$ is greater than 0.    
%
At line 24, we check if $i==1$, which means a match has been successful.
%
To find the appropriate operands, we need to right shift $x$ and $y$
such that the total number of their trailing zero blocks is $l-1$.
%
We add the matched $x*y$ to the match store $M$.
%
And, the algorithm proceeds for backtracking to find if more matching
exists.

\subsection{Matching Wallace tree multiplication}
\input{alg-match-wallace-tree}

A Wallace tree has a cascade of adders that take partial products 
and carry bits as input to produce the output bits.
%
In our matching algorithm, we find the set of inputs
to the adders for an output bit and classify them into
partial products and carry bits.
%
The half and full adders are defined as follows.
\begin{align*}
sumHalf(a,b) = a \lxor b  \qquad& \qquad sumFull(a,b,c) = a \lxor b \lxor c\\
carryHalf(a,b) = a \land b \qquad&\qquad
carryFull(a,b,c) = (a \land b) \lor (b \land c) \lor (c \land a).
\end{align*}
%
The sum output of a half/full adders are the result of 
xor operations of inputs.
%
To find the input to the adders, we start from an
output bit and follow backward until we find the input that
are not the result of some xor.

In Algorithm~\ref{alg:wallace}, we present a function
$\textsc{MatchWallace}$ that takes a $\qfbv$ term $t$ and returns a
set of matched multiplications.
%
At line 1, $t$ is matched with a concatenation of single bit terms $t_1$,..,$t_n$.
%
Similar to Algorithm~\ref{alg:long},
we maintain the partial product store $\Lambda$.
%
For each $i$,
we also maintain the multiset of terms $\Delta_i$ that were used as 
input to the adders for the $i$th bit.
%
In the loop at line 6, we traverse to the subterms until
a subterm is not the result of a xor.
%
In the traversal, we also collect the inputs of xors in $\Delta_i$, which
will help us in checking that all the carry inputs in adders for $t_{i+1}$
are generated by the adders for $t_i$.
%
If the term $t$ is not the result of xors then
we have the following possibilities.
%
\begin{itemize}
\item[line 10-13:]
  If $t$ is the carry bit of a half/full adder, and the inputs, the intermediate
  result of the sum bit, and the output sum bit
  of the adder are in $\Delta_{i-1}$. We remove the inputs and intermediate result
  of the adder from $\Delta_{i-1}$.
  We do not remove the output sum bit from $\Delta_{i-1}$, since it may be
  used as input to some other adder.
\item[line 14-15:] If $t$ is a partial product, we record it in $\Lambda_i$.
\item[line 16:] Otherwise, we return $\emptyset$.
\end{itemize}
At line 17, we check that $\Delta_{i-1} = \{t_{i-1}\}$, i.e., all carry bits from
the adders for $t_{i-1}$ are consumed by the adders for $t_i$ exactly once.
%
Again if the check fails then we return $\emptyset$.
%
After the loop at 3, we have collected the partial products in $\Lambda$.
%
At line 18, we call $\textsc{getMultOperands}(\Lambda,1)$ to get all
the matching multiplications.

\subsection{Our solver}
\input{alg-solver}

Using the above pattern matching algorithms, we modify an existing
solver $\textsc{SMTSolver}$, which
is presented in the Algorithm~\ref{alg:solver}.
%
$\textsc{OurSolver}$ adds the input formula $F$ in $\textsc{SMTSolver}$.
%
For every subterm of $F$, we attempt to match with both long multiplication
or Wallace tree multiplication.
%
For each discovered matching $x*y$, we add a bit-vector tautology $[x*y = t]$ 
to the solvers, which is obtained after
appropriately zero-padding $x$, $y$, and $t$.
%
\ashu{disjunctive assertion of the tautologies} 

% \ashu{Proof generation: Are we doing this??}

%--------------------- DO NOT ERASE BELOW THIS LINE --------------------------

%%% Local Variables:
%%% mode: latex
%%% TeX-master: "main"
%%% End:
